---
title: 'Assignment: An anlysis with NOAA Storm Database on severe weather events'
author: "Nanyuchentw"
date: "2023-10-07"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---
## Overview  
        1. Synopsis
        2. Data Processing  
                2.1 Load libraries  
                2.2 Read in data  
                2.3 Data transformation  
        3. Result

## 1. Synopsis  
This analysis used the NOAA Storm Database to find the event types that are either most harmful to population health or that caused the greatest economic damage. Data were first imported from the bz2 zipped file. The columns related to economic loss (PROPDMG, PROPDMGEXP, CROPDMG and CROPDMGEXP) were transformed from characters to numbers using the exponent indicated in the PROPDMGEXP and CROPDMGEXP columns. Total economic loss was then calculated by sum of property and crop loss amount. The fatality and injuries were calculated directed by the original data and grouped by event types. Finally, Top 5 most harmful event types were plotted.

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```
## 2. Data Processing
## 2.1 Load libraries
```{r, include=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
library(RColorBrewer)
```

## 2.2 Read in data
```{r}
# read in data
df <- read.csv("repdata_data_StormData.csv.bz2")
```

## 2.3 Data transformation
```{r}
#Because the PROPDMGEXP and CROPDMGEXP are exponents, 
#transform the PROPDMG and PROPDMGEXP columns to property loss amounts
df$PROPDMGEXP <- str_replace_all(df$PROPDMGEXP, "[Hh]", "2") 
df$PROPDMGEXP <- str_replace_all(df$PROPDMGEXP, "[Kk]", "3") 
df$PROPDMGEXP <- str_replace_all(df$PROPDMGEXP, "[Mm]", "6") 
df$PROPDMGEXP <- str_replace_all(df$PROPDMGEXP, "[Bb]", "9") 
df$PROPDMGEXP <- str_replace_all(df$PROPDMGEXP, "[?+-]", "0")
df$PROPDMGEXP[which(df$PROPDMGEXP=="")] <- "0"
# add a column for total property damage
df <- df %>% mutate(Prop_loss= as.numeric(PROPDMG)*(10^as.numeric(PROPDMGEXP)))

# transform the CROPDMG and CROPDMGEXP columns to crops loss amounts
df$CROPDMGEXP <- str_replace_all(df$CROPDMGEXP, "[Kk]", "3") 
df$CROPDMGEXP <- str_replace_all(df$CROPDMGEXP, "[Mm]", "6") 
df$CROPDMGEXP <- str_replace_all(df$CROPDMGEXP, "[Bb]", "9")
df$CROPDMGEXP <- str_replace_all(df$CROPDMGEXP, "[?]", "0")
df$CROPDMGEXP[which(df$CROPDMGEXP=="")] <- "0"
df <- df %>% mutate(Crop_loss= as.numeric(CROPDMG)*(10^as.numeric(CROPDMGEXP)))

#add a column for total economic loss
df <- df %>% mutate(Eco_loss= Crop_loss+ Prop_loss)

#find the type of events that cost the top 5 economic loss and then make a plot
E_cost <- df %>% group_by(EVTYPE) %>% 
        summarise(Total_loss= sum(Eco_loss), Crop_loss= sum(Crop_loss), Prop_loss= sum(Prop_loss)) %>% 
        arrange(desc(Total_loss)) %>% slice_head(n=10)

p <- E_cost %>% pivot_longer(Total_loss:Prop_loss, names_to = "Type", values_to = "Amount")
ggplot(p, aes(x=EVTYPE, y=Amount, fill=Type))+ geom_col(position="dodge")+
        scale_fill_brewer(palette = "Set1")+
        labs(x= "Event type", y="Economic loss (dollars)")+
        ggtitle(label= "Top 10 events with the greatest economic impact") +
        coord_flip()
```

#### The estimated amount of property and crop losses  
```{r}
kable(E_cost, "html") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

```{r}
#Evaluate event types that are most harmful to population health
#Find the event types that caused greatest injuries/fatalities
health_cost <- df %>% group_by(EVTYPE) %>% 
        summarise(injury= sum(INJURIES), death=sum(FATALITIES), total=sum(INJURIES+FATALITIES)) %>% 
        arrange(desc(total)) %>% slice_head(n=10)

p <- health_cost %>% pivot_longer(injury:total, names_to = "Type", values_to = "Number")
ggplot(p, aes(x= EVTYPE, y=Number, fill=Type))+ geom_col(position="dodge")+
        scale_fill_brewer(palette = "Set1")+
        labs(x="Event type", y="Injuries (person)")+
        ggtitle(label = "Top 10 events causing injuries/fatalities")+
        coord_flip()
```

#### The estimated injuries and fatalities  
```{r}
kable(health_cost, "html") %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

## 3. Result  
#### With respect to economic loss, Flood has the greatest economic consequences.  
#### To population health, Tornados are the most harmful event.  
